Testing for measurement invariance with respect to an ordinal variable

### 2012-07-27 ordinal/ordinal.Rnw
Additions to ordinal manuscript: o Modified dgp of simulation 1 to include 'constant' instability. o Started to add R code to the Sweave document. o Worked on data example for the manuscript.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@154 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\subsection{Method}
As described above, data were generated from a two-factor model
lacking measurement invariance in three of six unique variance
parameters.  Magnitude of measurement invariance violation, type of
measurement invariance violation, sample
size, and number of categories of the ordinal variable were manipulated.
We examined three sample sizes ($n=120, 480, 960$), three numbers of
categories ($m=4, 8, 12$), and
seven magnitudes of invariance 
violations.  The measurement invariance violations began
at level $1+l/2$ of $V$ and were constant thereafter
  The unique variances for
the ``violating'' levels deviated from the lower levels' unique
variances by $d$~times the parameters' asymptotic standard
errors (scaled by $\sqrt{n}$), with $d=0, 0.25, 0.5, \dots, 1.5$.

For each combination of $n \times m \times d$, 5,000 datasets were
generated and tested via the XX statistics described above.  In all
conditions, we maintained equal sample sizes at each level of the
ordinal variable.

%% sweave code here

@@

@@

>>>>

\subsection{Method}
As described above, data were generated from a two-factor model
lacking measurement invariance in three of six unique variance
parameters.  Magnitude of measurement invariance violation, type of
measurement invariance violation, sample
size, and number of categories of the ordinal variable were manipulated.
We examined three sample sizes ($n=120, 480, 960$), three numbers of
categories ($m=4, 8, 12$), and
seven magnitudes of invariance 
violations.  The measurement invariance violations began
at level $1+l/2$ of $V$ and were constant thereafter.
  The unique variances for
the ``violating'' levels deviated from the lower levels' unique
variances by $d$~times the parameters' asymptotic standard
errors (scaled by $\sqrt{n}$), with $d=0, 0.25, 0.5, \dots, 1.5$.

For each combination of $n \times m \times d$, 5,000 datasets were
generated and tested via the 4 statistics described above.  In all
conditions, we maintained equal sample sizes at each level of the
ordinal variable.

<<sim1>>=
## seed for replication
RNGkind(kind = "default", normal.kind = "default")
set.seed(1163)

if(file.exists("sim1.rda")) {
  load("sim1.rda")
} else {
  sim1 <- simulation()
  save(sim1, file="sim1.rda")
}
@

@@

Froh et al.\ \citeyear{FroFan11} used confirmatory factor analysis
models to study the invariance of three youth gratitude scales across
students aged 10 to 19 years.  Each of the scales were individually
factor-analyzed using the items comprising the scale.
For each model, the authors 
first fit a congeneric model (all parameters free for each age group),
followed by a tau-equivalent model (factor loadings restricted to be
equal across groups) and a parallel model (all parameters restricted
to be equal across groups).  Because their sample size was large ($n
\approx 1400$), they could not rely solely on likelihood ratio tests
(i.e., chi-square difference tests) for model comparison because the
tests were overly sensitive at their sample size.  To supplement these
tests, Froh et al examined a set of alternative fit indices, including
the non-normed fit index, the comparative fit index, and the
incremental fit index. %% TODO cite
The authors generally found support for the tau-equivalent models
through these alternative fit indices; the likelihood ratio test often
resulted in significance event when the alternative indices indicated
good fit.

In the analyses described below, we re-analyze the Froh et al.\ data using
the ordinal test statistics proposed in this paper.  This results in a
series of tests that are less sensitive to sample size than the
likelihood ratio test, while being
more sensitive to ``ordinal'' violations of measurement invariance.
We focus on two analyses from Froh et al.\ where the likelihood
ratio test resulted in significance (indicating that the restricted
model did not fit as well as the less-restricted model), but the
alternative fit measures indicated the opposite (using the rules of
thumb employed by Froh et al).  These include
comparison of a one-factor congeneric model to a one-factor
tau-equivalent model using the 
GQ6 and comparison of a one-factor tau-equivalent model to a one-factor
parallel model 
using the GAC.  To conduct equivalent analyses via the proposed tests, we
fit the restricted model and test for instability in the focal
model parameters.

Of the 1401 cases originally collected by Froh et al., we use here all
subjects with complete data (resulting in $n=1327$).

<<frohdata>>=
yg <- read.spss("grat data combined_9.29.09.sav", to.data.frame=TRUE)
names(yg) <- tolower(names(yg))

## Remove cases with 'imputed' values
yg <- yg[yg$ao_4 <= 9,]
decimals <- apply(yg[,4:28]%%1 > 0, 1, sum)
yg <- yg[decimals==0,]

## TODO Simulate critical values for ordL2BB
@

@@

The results section is divided into two subsections, one for each of
the analyses described previously.

<<gq6>>=
m1 <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
          data=yg, group="agegroup", group.equal="loadings",
          meanstructure=TRUE)
gefp1 <- gefp(m1, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=1:4)
omax1 <- sctest(gefp1, functional=ordwmax(gefp1))
ol2bb1 <- sctest(gefp1, functional=ordL2BB(gefp1))

## Full model
m1.f <- cfa('f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
            data=yg, group="agegroup", meanstructure=TRUE)
lrt1 <- anova(m1.f, m1)
@

\paragraph{GQ6.} Ordinal tests do not support LRT.

<<gac>>=
m2 <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal=c("loadings","residuals","lv.variances"), 
           meanstructure=TRUE)
gefp2 <- gefp(m2, fit = NULL, scores=estfun.lavaan, order.by = yg$agegroup, 
              vcov = info_full, sandwich = FALSE, parm=3:6)
omax2 <- sctest(gefp2, functional=ordwmax(gefp2))
ol2bb2 <- sctest(gefp2, functional=ordL2BB(gefp2))

## Full model
m2.f <- cfa('f1 =~ gac_1 + gac_2 + gac_3
           f1 ~ 0*1', data=yg, group="agegroup",
           group.equal="loadings", meanstructure=TRUE)
lrt2 <- anova(m2.f, m2)
@

\paragraph{GAC.} Ordinal tests do support LRT.
